\large \textbf{Burrows-Wheeler transform for terabases}

### 2015-10-28 paper/paper.tex
Polishing the paper.

\begin{abstract}
In order to avoid the reference bias introduced by mapping reads to a reference genome, bioinformaticians are investigating reference-free methods for analyzing sequenced genomes. With large projects sequencing thousands of individuals, this raises the need for tools capable of processing terabases of sequence data. A key method is the Burrows-Wheeler transform (BWT), which is widely used for compressing and indexing reads. We propose a practical algorithm for building the BWT of a large read collection by merging the BWTs of smaller subcollections. The algorithm is capable of merging over 700 Gbp/day on a single system, with 20--25 gigabytes of memory overhead on top of the run-length encoded BWTs.
\end{abstract}

@@

The \emph{Read Server project} at the Wellcome Trust Sanger Institute aims to avoid the reference bias entirely by developing tools for large-scale \emph{reference-free} genome analysis. Unique reads are compressed and indexed using the \emph{Burrows-Wheeler transform} (\BWT) \cite{Burrows1994}, while metadata databases contain information about the original reads. In the pilot phase, the Read Server project works with the low-coverage and exome data from phase 3 of the 1000 Genomes Project. After error correction and trimming the reads to either 73~bp or 100~bp, the 86~Tbp of raw data consisting of 922~billion original reads is reduced to 4.88~Tbp and 53.0~billion unique sequences. These sequences are stored in 16 \BWT-based indexes \cite{Ferragina2005a} taking a total of 561.5~gigabytes.

@@

\begin{table}[t!]
\begin{center}
\caption{Datasets used in the experiments.}\label{table:datasets}\smallskip%
{
\renewcommand{\baselinestretch}{1}\footnotesize
\begin{tabular}{cc|ccc}
\hline
\textbf{Dataset} & \textbf{File} & \textbf{Size} & \textbf{Reads} & \textbf{Read length} \\
\hline
\CEU & NA12878 &  284 Gbp & 2.81G &       101 bp \\
     & NA12891 &  242 Gbp & 2.40G &       101 bp \\
     & NA12892 &  245 Gbp & 2.42G &       101 bp \\
     & Merged  &  771 Gbp & 7.63G &       101 bp \\
\hline
\RS  & AA      &  433 Gbp & 4.69G & 73 or 100 bp \\
     & TT      &  432 Gbp & 4.68G & 73 or 100 bp \\
     & AT      &  275 Gbp & 2.98G & 73 or 100 bp \\
     & TA      &  355 Gbp & 3.84G & 73 or 100 bp \\
     & Merged  & 1.49 Tbp & 16.2G & 73 or 100 bp \\
\hline
\RS  & *A, *C  & 2.45 Tbp & 26.5G & 73 or 100 bp \\
     & *G, *T  & 2.44 Tbp & 26.5G & 73 or 100 bp \\
\hline
\end{tabular}}
\end{center}
\end{table}

@@

\item \RS{} is from the \emph{Read Server project}, which uses all low-coverage and exome data from the phase 3. After error correction, trimming the reads to 73 or 100~bp, and merging the duplicates, there are 53.0~billion unique reads for a total of 4.88~Tbp. The reads are in 16 run-length encoded \BWT{}s built using the \emph{String Graph Assembler} \cite{Simpson2012}, distributed according to the last two bases.

@@

\smallbreak\noindent\textbf{Benchmarking.} For benchmarking with different parameter values, we converted four \BWT{} files (AA, TT, AT, and TA) containing a total of 1.49~Tbp from the \RS{} dataset to the \emph{native format} of \BWTmerge. This format includes the \BWT{} and the \rank/\select{} structures required by the FM-index. We then merged the \BWT{}s (in the given order). We used 128~MB or 256~MB run buffers and 256~MB or 512~MB thread buffers. The number of merge buffers was 4 or 5 with 512~MB thread buffers and 5 or 6 with 256~MB thread buffers, so that the files on disk were always merged from either 8~GB or 16~GB of thread buffers.

\begin{figure}[t!]
\begin{center}
\includegraphics{benchmark.pdf}%
\hspace{-0.3in}%
\includegraphics{comparison.pdf}
\end{center}
\caption{Left: Time/space trade-offs for merging four \BWT{} files containing a total of 1.49~Tbp. Label rXbYmZ denotes $X$~MB run buffers, $Y$~MB threads buffers, and $Z$ merge buffers. The dashed line represents the total size of the last pair of \BWT{}s to be merged. Right: Time/space trade-offs for building the \BWT{} of the 771~Gbp \CEU{} dataset.}\label{fig:benchmark}
\end{figure}

@@

For the further experiments, we chose 128~MB run buffers and 256~MB thread buffers. We use 5 merge buffers in the \emph{small} variant (memory overhead 21.1~GB) and 6 merge buffers in the \emph{fast} variant (overhead 30.8~GB).

\smallbreak\noindent\textbf{Comparison.} In the next experiment, we compared the performance of \BWTmerge{} to the \ropebwt{} \cite{Li2011-2013} implementation of the BCR algorithm \cite{Bauer2013} and to \ropebwtii{} \cite{Li2014a}. These two tools are the fastest existing \BWT{} construction tools for read collections on general hardware \cite{Li2014a}. We ran \ropebwt{} with parameters \texttt{-btORf -abcr} and \ropebwtii{} with parameters \texttt{-bRm10g}. Due to memory constraints, we used the smaller \CEU{} dataset.

We built the \BWT{} of the dataset in four ways. We concatenated all input files, and built the \BWT{} directly for the entire collection using both \ropebwt{} and \ropebwtii{}. We built the \BWT{} for each input file separately using \ropebwt{}, and then merged the results using \BWTmerge{} with small settings. Finally, we built the individual \BWT{}s using parallel \ropebwt{} jobs, and then merged them using \BWTmerge{} with fast settings. In each case, the tools were allowed to output the \BWT{} files in their preferred formats.

The results can be seen in Figure~\ref{fig:benchmark} (right).
% FIXME discussion of the results; note that ropebwt2 did not finish

\smallbreak\noindent\textbf{Read Server.} In the final experiment, we merged the 16 \BWT{} files in the \RS{} dataset into two files using the fast settings.
% FIXME sizes in native/sga format, merging time/space

@@

\BWTmerge{} can be used as a building block of a \BWT{} construction algorithm. We split the read collection into multiple subcollections, build the \BWT{}s of the subcollections with other tools, and merge the results. Depending on whether we run one or more \BWT{} construction jobs in parallel, the resulting algorithm can be slower but more space-efficient than the existing algorithms, or faster than them with similar memory usage.

The most important feature of our algorithm is its low memory usage. With it, we can build the \BWT{}s of much larger read collections than before on commonly available hardware. As a concrete example, we merged the 16 Read Server \BWT{} files into two files. This reduced the number of servers required to house the indexes from three to two, and also improved the query performance of the servers.

@@

Read Server stores the reads in \emph{reverse lexicographic order} to improve compression \cite{Cox2012}. We can maintain this order with a few changes to the search phase \cite{Li2014a}. Sorting the reads by their \emph{reverse complements} also improves compression in a similar way. In this order, $\mSA[\mBWT.\mselect(i,0)]$ points to the beginning of the reverse complement of read $T_{i}$, if the collection includes the reverse complement of each read \cite{Li2014a}.

>>>>

\begin{abstract}
In order to avoid the reference bias introduced by mapping reads to a reference genome, bioinformaticians are investigating reference-free methods for analyzing sequenced genomes. With large projects sequencing thousands of individuals, this raises the need for tools capable of processing terabases of sequence data. A key method is the Burrows-Wheeler transform (BWT), which is widely used for compressing and indexing reads. We propose a practical algorithm for building the BWT of a large read collection by merging the BWTs of smaller subcollections. The algorithm is capable of merging over 700 Gbp/day on a single system, with 20--30 gigabytes of memory overhead on top of the run-length encoded BWTs.
\end{abstract}

@@

The \emph{Read Server} project at the Wellcome Trust Sanger Institute aims to avoid the reference bias entirely by developing tools for large-scale \emph{reference-free} genome analysis. Unique reads are compressed and indexed using the \emph{Burrows-Wheeler transform} (\BWT) \cite{Burrows1994}, while metadata databases contain information about the original reads. In the pilot phase, the Read Server project works with the low-coverage and exome data from phase 3 of the 1000 Genomes Project. After error correction and trimming the reads to either 73~bp or 100~bp, the 86~Tbp of raw data consisting of 922~billion original reads is reduced to 4.88~Tbp and 53.0~billion unique sequences. These sequences are stored in 16 \BWT-based indexes \cite{Ferragina2005a} taking a total of 561.5~gigabytes.

@@

\begin{table}[t!]
\begin{center}
\caption{Datasets used in the experiments. \BWT{} sizes are for the native format. RLO indicates that the reads are sorted in reverse lexicographic order in each input file.}\label{table:datasets}\smallskip%
{
\renewcommand{\baselinestretch}{1}\footnotesize
\begin{tabular}{cc|ccc|cc}
\hline
 & & & & & \multicolumn{2}{c}{\textbf{\BWT{} size}} \\
\textbf{Dataset} & \textbf{File} & \textbf{Size} & \textbf{Reads} & \textbf{Read length} & \textbf{Unsorted} & \textbf{RLO} \\
\hline
\CEU & NA12878 &  284 Gbp & 2.81G &       101 bp & 50.3 GB \\
     & NA12891 &  242 Gbp & 2.40G &       101 bp & 42.4 GB \\
     & NA12892 &  245 Gbp & 2.42G &       101 bp & 43.8 GB \\
     & Merged  &  771 Gbp & 7.63G &       101 bp & \\
\hline
\RS  & AA      &  433 Gbp & 4.69G & 73 or 100 bp &      -- & 38.5 GB \\
     & TT      &  432 Gbp & 4.68G & 73 or 100 bp &      -- & 38.7 GB \\
     & AT      &  275 Gbp & 2.98G & 73 or 100 bp &      -- & 26.6 GB \\
     & TA      &  355 Gbp & 3.84G & 73 or 100 bp &      -- & 32.7 GB \\
     & Merged  & 1.49 Tbp & 16.2G & 73 or 100 bp &      -- &  117 GB \\
\hline
\RS  & *A, *C  & 2.45 Tbp & 26.5G & 73 or 100 bp &      -- &  225 GB \\
     & Merged  & 2.45 Tbp & 26.5G & 73 or 100 bp &      -- & \\
\hline
\RS  & *G, *T  & 2.44 Tbp & 26.5G & 73 or 100 bp &      -- &  226 GB \\
     & Merged  & 2.44 Tbp & 26.5G & 73 or 100 bp &      -- & \\
\hline
\end{tabular}}
\end{center}
\end{table}

@@

\item \RS{} is from the \emph{Read Server} project, which uses all low-coverage and exome data from the phase 3. After error correction, trimming the reads to 73 or 100~bp, and merging the duplicates, there are 53.0~billion unique reads for a total of 4.88~Tbp. The reads are in 16 run-length encoded \BWT{}s built using the \emph{String Graph Assembler} \cite{Simpson2012}, distributed according to the last two bases.

@@

\smallbreak\noindent\textbf{Parameters.} For benchmarking with different parameter values, we converted four \BWT{} files (AA, TT, AT, and TA) containing a total of 1.49~Tbp from the \RS{} dataset to the \emph{native format} of \BWTmerge. This format includes the \BWT{} and the \rank/\select{} structures required by the FM-index. We then merged the \BWT{}s (in the given order). We used 128~MB or 256~MB run buffers and 256~MB or 512~MB thread buffers. The number of merge buffers was 4 or 5 with 512~MB thread buffers and 5 or 6 with 256~MB thread buffers, so that the files on disk were always merged from either 8~GB or 16~GB of thread buffers.

\begin{figure}[t!]
\begin{center}
\includegraphics{benchmark.pdf}%
\hspace{-0.3in}%
\includegraphics{comparison.pdf}
\end{center}
\caption{Left: Time/space trade-offs for merging four \BWT{} files containing a total of 1.49~Tbp. Label rXbYmZ denotes $X$~MB run buffers, $Y$~MB threads buffers, and $Z$ merge buffers. The dashed line represents the total size of the last pair of \BWT{}s to be merged. Right: Time/space trade-offs for building the \BWT{} of the 771~Gbp \CEU{} dataset. RLO indicates reverse lexicographic order.}\label{fig:benchmark}
\end{figure}

@@

For the further experiments, we chose 128~MB run buffers, and 256~MB thread buffers and 6 merge buffers (overhead 30.8~GB).

\smallbreak\noindent\textbf{Comparison.} In the next experiment, we compared the performance of \BWTmerge{} to the \ropebwt{} \cite{Li2011-2013} implementation of the BCR algorithm \cite{Bauer2013} and to \ropebwtii{} \cite{Li2014a}. These two tools are the fastest existing \BWT{} construction tools for read collections on general hardware \cite{Li2014a}. We ran \ropebwt{} with parameters \texttt{-btORf -abcr} and \ropebwtii{} with parameters \texttt{-bRm10g}. For some runs, we also used \texttt{-s} with both tools to sort the reads in \emph{reverse lexicographic order} (RLO) to improve compression. Due to memory constraints, we used the smaller \CEU{} dataset.

We built the \BWT{} of the dataset in several ways. We concatenated all input files, and built the \BWT{} directly for the entire collection using both \ropebwt{} and \ropebwtii{}. We also built the \BWT{} for each input file separately using \ropebwt{}, and then merged the results using \BWTmerge{}. Finally, we repeated everything with the reads sorted in reverse lexicographic order. In each case, the tools were allowed to output the \BWT{} files in their preferred formats.

The results can be seen in Figure~\ref{fig:benchmark} (right).
% FIXME discussion of the results; note that ropebwt2 did not finish without RLO; BWT-merge does not do RLO

\smallbreak\noindent\textbf{Read Server.} In the final experiment, we merged the 16 \BWT{} files in the \RS{} dataset into two files.
% FIXME sizes in native/sga format, merging time/space

@@

\BWTmerge{} can be used as a building block of a \BWT{} construction algorithm. We split the read collection into multiple subcollections, build the \BWT{}s of the subcollections with other tools, and merge the results. The resulting algorithm is typically slower but more space-efficient than the existing algorithms.

The most important feature of our algorithm is its low memory usage. With it, we can build the \BWT{}s of much larger read collections than before on commonly available hardware. As a concrete example, we merged the 16 Read Server \BWT{} files into two files. This reduced the number of servers required to use the indexes from three to two, and also improved the query performance of the servers.

@@

Read Server stores the reads in \emph{reverse lexicographic order} to improve compression \cite{Cox2012}. We can maintain this order with a few changes to the search phase \cite{Li2014a}. Sorting the reads by their \emph{reverse complements} also improves compression in a similar way. In this order, $\mSA[\mBWT.\mselect(i,0)]$ points to the beginning of the reverse complement of read $T_{i}$, if the collection includes the reverse complement of each read \cite{Li2014a}. With \emph{lexicographic order}, we can determine $\mSA[i]$ without samples by using $\mLF$ instead of $\Psi$, which is often faster in practice.
